---
title: "Exploratory Data Analysis for Point-referenced (Geostatistical) Data"
output: html_document
date: "2025-09-18"
---

We will use some R libraries to carry out some exploratory data analysis for point-referenced data.

1. Required libraries
```{r}
library(spBayes)
library(classInt)
library(RColorBrewer)
```

2. Load the Western Experimental Forest (WEF) dataset from the "spBayes" package. Extract the rows that do not contain missing values.
```{r}
data(WEF.dat)
WEF.dat <- WEF.dat[!apply(WEF.dat[,c("East_m","North_m", "DBH_cm","Tree_height_m","ELEV_m")], 1, function(x)any(is.na(x))),]
```

3. We will take a closer look at the variable "Diameter at Breast Height" (DBH). We will plot the locations as circles where DBH is measured and indicate high and low values of DBH by the size of the circles.   
```{r}
DBH <- WEF.dat$DBH_cm
HT <- WEF.dat$Tree_height_m
coords <- as.matrix(WEF.dat[,c("East_m","North_m")])
plot(coords, pch=1, cex=sqrt(DBH)/10, col="darkgreen", xlab="Easting (m)", ylab="Northing (m)")
leg.vals <- round(quantile(DBH),0)
legend("topleft", pch=1, legend=leg.vals, col="darkgreen",
pt.cex=sqrt(leg.vals)/10, bty="n", title="DBH (cm)")
```

4. Create a color palette for subsequent plots. We do this by creating intervals based on the range of values of DBH and assigning colors to each interval.   
```{r}   
col.br <- colorRampPalette(c("blue", "cyan", "yellow", "red"))
col.pal <- col.br(5)
fixed <- classIntervals(DBH, n = 4, style = "fixed", fixedBreaks = c(0, 12.7, 30.48, 60, max(DBH) + 1))
fixed.col <- findColours(fixed, col.pal)
plot(coords, col = fixed.col, pch = 19, cex = 0.5, main = "Forestry tree size classes", xlab = "Easting (m)", ylab = "Northing (m)")
legend("topleft", fill = attr(fixed.col, "palette"), legend = c("sapling", "poletimber", "sawtimber", "large sawtimber"), bty = "n")
```

5. Load the MBA and fields libraries for creating surface interpolation plots
```{r}
library(MBA)
library(fields) ## For using the image.plot function
x.res <- 100
y.res <- 100
surf <- mba.surf(cbind(coords, DBH), no.X = x.res, no.Y = y.res, h = 5, m = 2, extend = FALSE)$xyz.est
image.plot(surf, xaxs = "r", yaxs = "r", xlab = "Easting (m)", ylab = "Northing (m)", col = col.br(25))
contour(surf, add=T) ## (Optional) Adds contour lines to the plot
```

6. Load the "geoR" library for computing variograms.
```{r}   
library(geoR)
max.dist <- 0.25 * max(iDist(coords))
bins <- 50

vario.DBH <- variog(coords = coords, data = DBH, uvec = (seq(0, max.dist, length = bins)))
fit.DBH <- variofit(vario.DBH, ini.cov.pars = c(600,200/-log(0.05)), cov.model = "exponential", minimisation.function = "nls", weights = "equal")
```

7. Run an Ordinary Least Squares (OLS) regression and compute the variograms from the residuals.
```{r}   
lm.DBH <- lm(DBH ~ Species, data = WEF.dat)
summary(lm.DBH)
DBH.resid <- resid(lm.DBH)

vario.DBH.resid <- variog(coords = coords, data = DBH.resid, uvec = (seq(0, max.dist, length = bins)))
fit.DBH.resid <- variofit(vario.DBH.resid, ini.cov.pars = c(300, 200/-log(0.05)), cov.model = "exponential", minimisation.function = "nls", weights = "equal")

par(mfrow = c(1, 2))
plot(vario.DBH, ylim = c(200, 1200), main = "DBH")
lines(fit.DBH)
abline(h = fit.DBH$nugget, col = "blue")
abline(h = fit.DBH$cov.pars[1] + fit.DBH$nugget, col = "green")
abline(v = -log(0.05) * fit.DBH$cov.pars[2], col = "red3")
plot(vario.DBH.resid, ylim = c(200, 500), main = "DBH residuals")
lines(fit.DBH.resid)
abline(h = fit.DBH.resid$nugget, col = "blue")
abline(h = fit.DBH.resid$cov.pars[1] + fit.DBH.resid$nugget, col = "green")
abline(v = -log(0.05) * fit.DBH.resid$cov.pars[2], col = "red3")
```

8. Illustrate variograms using the "gstat" package
```{r}   
library(sp)
library(gstat)
ELEV_m <- WEF.dat$ELEV_m
sp.dat <- as.data.frame(cbind(DBH, HT, coords, ELEV_m))
coordinates(sp.dat) <- c("East_m", "North_m")
vario.DBH <- variogram(DBH ~ ELEV_m, data = sp.dat,utoff = max.dist, width = 5, alpha = (0:3) * 45)
fit.DBH <- fit.variogram(vario.DBH, vgm(1000, "Exp", 200/-log(0.05), 600))
print(plot(vario.DBH, fit.DBH))
```

