---
title: "Exact Federated Learning Inference in Bayesian Multivariate Linear Regression"
author: "Sudipto Banerjee"
output: 
  html_document:
    toc: true
    number_sections: true
---

*Note:* To generate this HTML file, run `renderExactMvLM.sh` in the terminal from the `src/bigDataRegression/mniw/exactMvLM/run` directory. Alternatively find the source `Rmd` and `R` files in that same folder starting from the `BGC_2023` GitHub home.

```bash
# Navigate to the directory and run the compilation script
cd src/bigDataRegression/mniw/exactMvLM/run
./renderExactMvLM.sh
```


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(knitr)
library(mniw)
source("../src/mniwSamplerFederated.r")
```

# Conjugate Bayesian Multivariate Linear Regression

A conjugate Bayesian multivariate linear model (the MNIW model) is given by:

$$Y \mid \beta, \Sigma, X \sim MN_{n,q}(X\beta, I_n, \Sigma)$$
$$\beta \mid \Sigma \sim MN_{p,q}(M_0 m_0, M_0, \Sigma)$$
$$\Sigma \sim IW(\nu_0, \Psi_0)$$

Unknown parameters include the regression coefficient matrix and the covariance matrix, i.e., $\theta = \{\beta, \Sigma\}$.

The likelihood of the data is:

$$p(Y \mid \beta, \Sigma) \propto |\Sigma|^{-n/2} \exp\left\{ -\frac{1}{2} \text{tr}\left[ \Sigma^{-1} (Y - X\beta)^{\top} (Y - X\beta) \right] \right\}$$

The joint posterior density is $p(\beta, \Sigma \mid Y) \propto p(\Sigma) \times p(\beta \mid \Sigma) \times p(Y \mid \beta, \Sigma)$.

---

# Deriving the Joint Posterior $p(\beta, \Sigma \mid Y)$

$$\begin{aligned}
p(\beta, \Sigma \mid Y) &\propto IW(\Sigma \mid \nu_0, \Psi_0) \times MN(\beta \mid M_0 m_0, M_0, \Sigma) \times MN(Y \mid X\beta, I_n, \Sigma) \\
&\propto |\Sigma|^{-\frac{\nu_0 + q + 1}{2}} \exp\left\{ -\frac{1}{2} \text{tr}\left( \Sigma^{-1} \Psi_0 \right) \right\} \\
&\quad \times |\Sigma|^{-\frac{p}{2}} \exp\left\{ -\frac{1}{2} \text{tr}\left[ \Sigma^{-1} (\beta - M_0 m_0)^{\top} M_0^{-1} (\beta - M_0 m_0) \right] \right\} \\
&\quad \times |\Sigma|^{-\frac{n}{2}} \exp\left\{ -\frac{1}{2} \text{tr}\left[ \Sigma^{-1} (Y - X\beta)^{\top} (Y - X\beta) \right] \right\} \\
&\propto |\Sigma|^{-\frac{\nu_0 + n + p + q + 1}{2}} \exp\left\{ -\frac{1}{2} \text{tr}\left( \Sigma^{-1} \left[ \Psi_0 + Q(\beta, Y) \right] \right) \right\}
\end{aligned}$$

where the quadratic terms in $\beta$ are combined as:

$$Q(\beta, Y) = (\beta - M_0 m_0)^{\top} M_0^{-1} (\beta - M_0 m_0) + (Y - X\beta)^{\top} (Y - X\beta)$$

---

# Matrix Completion of Square

Expanding and recombining $Q(\beta, Y)$ (noting that $M_0$ is symmetric, so $(M_0 m_0)^\top M_0^{-1} \beta = m_0^\top M_0 M_0^{-1} \beta = m_0^\top \beta$):

$$\begin{aligned}
Q(\beta, Y) &= (\beta - M_0 m_0)^{\top} M_0^{-1} (\beta - M_0 m_0) + (Y - X\beta)^{\top} (Y - X\beta) \\
&= \beta^{\top} M_0^{-1} \beta - \beta^{\top} m_0 - m_0^{\top} \beta + m_0^{\top} M_0 m_0 + Y^{\top} Y - Y^{\top} X \beta - \beta^{\top} X^{\top} Y + \beta^{\top} X^{\top} X \beta \\
&= \beta^{\top} (M_0^{-1} + X^{\top} X) \beta - \beta^{\top} (m_0 + X^{\top} Y) - (m_0^{\top} + Y^{\top} X) \beta + (m_0^{\top} M_0 m_0 + Y^{\top} Y) \\
&= (\beta - Mm)^{\top} M^{-1} (\beta - Mm) + S^\ast
\end{aligned}$$

where the updated parameters $M$, $m$, and $S^\ast$ are defined as:

$$M^{-1} = M_0^{-1} + X^{\top} X, \quad m = m_0 + X^{\top} Y, \quad S^\ast = m_0^{\top} M_0 m_0 + Y^{\top} Y - m^{\top} M m$$

---

# Factorizing the Joint Posterior

Substituting the completed square back into the joint posterior yields:

$$p(\beta, \Sigma \mid Y) \propto \left( |\Sigma|^{-\frac{\nu + q + 1}{2}} \exp\left\{ -\frac{1}{2} \text{tr}\left[ \Sigma^{-1} \Psi \right] \right\} \right) \times \left( |\Sigma|^{-\frac{p}{2}} \exp\left\{ -\frac{1}{2} \text{tr}\left[ \Sigma^{-1} (\beta - Mm)^{\top} M^{-1} (\beta - Mm) \right] \right\} \right)$$

where:

$$\nu = \nu_0 + n, \quad \Psi = \Psi_0 + S^\ast = \Psi_0 + m_0^{\top} M_0 m_0 + Y^{\top} Y - m^{\top} M m$$

This factorizes into $IW(\Sigma \mid \nu, \Psi) \times MN(\beta \mid Mm, M, \Sigma)$.

---

# Federated Learning: Decentralized Divide & Conquer

In a Federated Learning architecture, a central server orchestrates statistical estimation across $T$ local client servers. Raw client datasets $(Y_t, X_t)$ remain strictly on client $t$ to ensure data privacy.

The central server holds global prior parameters: $M_0^{-1}$, $m_0$, $\nu_0$, and $\Psi_0$.

## Local Client Computations

Each client $t$ independently computes localized sufficient statistics:

$$\Delta M_t^{-1} = X_t^{\top} X_t, \quad \Delta m_t = X_t^{\top} Y_t, \quad \Delta Q_t = Y_t^{\top} Y_t, \quad n_t = \text{nrow}(X_t)$$

Client $t$ transmits only the low-dimensional tuple $(\Delta M_t^{-1}, \Delta m_t, \Delta Q_t, n_t)$ back to the central server. For $p=10$ predictors and $q=5$ outcomes, the client transmits a $10 \times 10$ matrix, a $10 \times 5$ matrix, a $5 \times 5$ matrix, and one scalar.

## Server One-Shot Aggregation

The server aggregates client statistics in a single pass:

$$M_T^{-1} = M_0^{-1} + \sum_{t=1}^T \Delta M_t^{-1}, \quad m_T = m_0 + \sum_{t=1}^T \Delta m_t, \quad \nu_T = \nu_0 + \sum_{t=1}^T n_t$$

Summing across all $T$ clients telescopes to yield the global scale matrix $\Psi_T$:

$$\Psi_T = \Psi_0 + \sum_{t=1}^T \Delta Q_t + m_0^{\top} M_0 m_0 - m_T^{\top} M_T m_T$$

The central server draws global posterior samples directly from $IW(\nu_T, \Psi_T)$ and $MN(M_T m_T, M_T, \Sigma)$.

---

# Empirical Analysis: Custom Federated Sampler (`mniwSamplerFederated`)

Loading data from `../data/X.txt` and `../data/Y.txt`, partitioning across $T = 5$ federated clients, and comparing single-pass federated estimation against full single-server batch updates.

```{r execute_federated_custom}
# Ingest dataset files
X_raw  <- as.matrix(read.table("../data/X.txt"))
Y_raw  <- as.matrix(read.table("../data/Y.txt"))
X_full <- cbind(1, X_raw)
Y_full <- Y_raw

p_dim <- ncol(X_full)
q_dim <- ncol(Y_full)

# Global prior parameters
m0     <- matrix(0, nrow = p_dim, ncol = q_dim)
M0_inv <- diag(1, p_dim)
nu0    <- q_dim + 2
Psi0   <- diag(1, q_dim)

# Partition data across T = 5 clients
T_clients   <- 5
client_idxs <- split(seq_len(nrow(X_full)), cut(seq_len(nrow(X_full)), T_clients, labels = FALSE))
client_data_list <- lapply(client_idxs, function(idx) {
  list(X = X_full[idx, ], Y = Y_full[idx, ])
})

# 1. Federated Learning Sampler (Custom R)
set.seed(42)
res_fed_custom <- mniwSamplerFederated(client_data_list, m0, M0_inv, nu0, Psi0, n_samples = 2000)

# 2. Single-Server Full Batch Sampler
full_server_data <- list(list(X = X_full, Y = Y_full))
set.seed(42)
res_batch_custom <- mniwSamplerFederated(full_server_data, m0, M0_inv, nu0, Psi0, n_samples = 2000)
```

## Parameter Precision Comparison (Custom Sampler)

```{r compare_parameters_custom}
diff_m     <- max(abs(res_fed_custom$post_params$m - res_batch_custom$post_params$m))
diff_M_inv <- max(abs(res_fed_custom$post_params$M_inv - res_batch_custom$post_params$M_inv))
diff_Psi   <- max(abs(res_fed_custom$post_params$Psi - res_batch_custom$post_params$Psi))
diff_nu    <- abs(res_fed_custom$post_params$nu - res_batch_custom$post_params$nu)

param_comp <- data.frame(
  Parameter_Matrix = c("Location Vector (m)", "Precision Matrix (M^{-1})", "Scale Matrix (Psi)", "Degrees of Freedom (nu)"),
  Max_Absolute_Difference = c(diff_m, diff_M_inv, diff_Psi, diff_nu)
)

kable(param_comp, digits = 16, caption = "Numerical Equivalence: Federated vs. Single Server (Custom Sampler)")
```

## Posterior Parameter Summaries (Custom Sampler)

```{r posterior_summary_custom, echo=FALSE, results='asis'}
calc_summaries <- function(arr, name) {
  grid <- expand.grid(Row = 1:dim(arr)[1], Col = 1:dim(arr)[2])
  res <- lapply(seq_len(nrow(grid)), function(k) {
    i <- grid$Row[k]; j <- grid$Col[k]
    vals <- arr[i, j, ]
    data.frame(
      Param = paste0(name, "[", i, ",", j, "]"),
      Mean = mean(vals), SD = sd(vals),
      `2.5%` = quantile(vals, 0.025), `97.5%` = quantile(vals, 0.975),
      check.names = FALSE
    )
  })
  do.call(rbind, res)
}

df_beta_fed  <- calc_summaries(res_fed_custom$samples$Beta, "Beta")
df_sigma_fed <- calc_summaries(res_fed_custom$samples$Sigma, "Sigma")

cat("### Regression Coefficients (Beta) Summary (Custom Federated Sampler)\n\n")
print(kable(df_beta_fed, digits = 4, row.names = FALSE))

cat("\n### Covariance Matrix (Sigma) Summary (Custom Federated Sampler)\n\n")
print(kable(df_sigma_fed, digits = 4, row.names = FALSE))
```

## Quantile-Quantile (QQ) Validation (Custom Sampler)

### Regression Coefficients ($\beta$) QQ-Plots

```{r qqplot_beta_custom, fig.width=9, fig.height=7, echo=FALSE}
par(mfrow = c(p_dim, q_dim), mar = c(3, 3, 2, 1))
grid_beta <- expand.grid(Row = 1:p_dim, Col = 1:q_dim)

invisible(lapply(seq_len(nrow(grid_beta)), function(k) {
  i <- grid_beta$Row[k]; j <- grid_beta$Col[k]
  qqplot(res_fed_custom$samples$Beta[i, j, ], res_batch_custom$samples$Beta[i, j, ],
         main = paste0("Beta[", i, ",", j, "]"),
         xlab = "Federated Quantiles", ylab = "Single Server Quantiles",
         pch = 16, col = rgb(0.2, 0.4, 0.8, 0.4))
  abline(0, 1, col = "firebrick", lwd = 2)
}))
par(mfrow = c(1, 1))
```

### Covariance Matrix ($\Sigma$) QQ-Plots

```{r qqplot_sigma_custom, fig.width=8, fig.height=8, echo=FALSE}
par(mfrow = c(q_dim, q_dim), mar = c(3, 3, 2, 1))
grid_sigma <- expand.grid(Row = 1:q_dim, Col = 1:q_dim)

invisible(lapply(seq_len(nrow(grid_sigma)), function(k) {
  i <- grid_sigma$Row[k]; j <- grid_sigma$Col[k]
  qqplot(res_fed_custom$samples$Sigma[i, j, ], res_batch_custom$samples$Sigma[i, j, ],
         main = paste0("Sigma[", i, ",", j, "]"),
         xlab = "Federated Quantiles", ylab = "Single Server Quantiles",
         pch = 16, col = rgb(0.8, 0.4, 0.2, 0.4))
  abline(0, 1, col = "firebrick", lwd = 2)
}))
par(mfrow = c(1, 1))
```

---

# Empirical Analysis: C++ Accelerated Federated Sampler (`mniwSamplerFederated_rmniw`)

Running federated estimation with `mniwSamplerFederated_rmniw`, passing precision $\Omega = M_T^{-1}$ directly to C++ accelerated `mniw::rmniw`.

```{r execute_federated_rmniw}
# 1. Federated Learning Sampler (rmniw)
set.seed(42)
res_fed_rmniw <- mniwSamplerFederated_rmniw(client_data_list, m0, M0_inv, nu0, Psi0, n_samples = 2000)

# 2. Single-Server Full Batch Sampler (rmniw)
set.seed(42)
res_batch_rmniw <- mniwSamplerFederated_rmniw(full_server_data, m0, M0_inv, nu0, Psi0, n_samples = 2000)
```

## Parameter Precision Comparison (`rmniw` Sampler)

```{r compare_parameters_rmniw}
diff_m_rmniw     <- max(abs(res_fed_rmniw$post_params$m - res_batch_rmniw$post_params$m))
diff_M_inv_rmniw <- max(abs(res_fed_rmniw$post_params$M_inv - res_batch_rmniw$post_params$M_inv))
diff_Psi_rmniw   <- max(abs(res_fed_rmniw$post_params$Psi - res_batch_rmniw$post_params$Psi))
diff_nu_rmniw    <- abs(res_fed_rmniw$post_params$nu - res_batch_rmniw$post_params$nu)

param_comp_rmniw <- data.frame(
  Parameter_Matrix = c("Location Vector (m)", "Precision Matrix (M^{-1})", "Scale Matrix (Psi)", "Degrees of Freedom (nu)"),
  Max_Absolute_Difference = c(diff_m_rmniw, diff_M_inv_rmniw, diff_Psi_rmniw, diff_nu_rmniw)
)

kable(param_comp_rmniw, digits = 16, caption = "Numerical Equivalence: Federated vs. Single Server (rmniw Sampler)")
```

## Posterior Parameter Summaries (`rmniw` Sampler)

```{r posterior_summary_rmniw, echo=FALSE, results='asis'}
df_beta_fed_rmniw  <- calc_summaries(res_fed_rmniw$samples$Beta, "Beta")
df_sigma_fed_rmniw <- calc_summaries(res_fed_rmniw$samples$Sigma, "Sigma")

cat("### Regression Coefficients (Beta) Summary (rmniw Federated Sampler)\n\n")
print(kable(df_beta_fed_rmniw, digits = 4, row.names = FALSE))

cat("\n### Covariance Matrix (Sigma) Summary (rmniw Federated Sampler)\n\n")
print(kable(df_sigma_fed_rmniw, digits = 4, row.names = FALSE))
```

## Quantile-Quantile (QQ) Validation (`rmniw` Sampler)

### Regression Coefficients ($\beta$) QQ-Plots (`rmniw`)

```{r qqplot_beta_rmniw, fig.width=9, fig.height=7, echo=FALSE}
par(mfrow = c(p_dim, q_dim), mar = c(3, 3, 2, 1))

invisible(lapply(seq_len(nrow(grid_beta)), function(k) {
  i <- grid_beta$Row[k]; j <- grid_beta$Col[k]
  qqplot(res_fed_rmniw$samples$Beta[i, j, ], res_batch_rmniw$samples$Beta[i, j, ],
         main = paste0("Beta[", i, ",", j, "]"),
         xlab = "Federated rmniw Quantiles", ylab = "Single Server rmniw Quantiles",
         pch = 16, col = rgb(0.2, 0.4, 0.8, 0.4))
  abline(0, 1, col = "firebrick", lwd = 2)
}))
par(mfrow = c(1, 1))
```

### Covariance Matrix ($\Sigma$) QQ-Plots (`rmniw`)

```{r qqplot_sigma_rmniw, fig.width=8, fig.height=8, echo=FALSE}
par(mfrow = c(q_dim, q_dim), mar = c(3, 3, 2, 1))

invisible(lapply(seq_len(nrow(grid_sigma)), function(k) {
  i <- grid_sigma$Row[k]; j <- grid_sigma$Col[k]
  qqplot(res_fed_rmniw$samples$Sigma[i, j, ], res_batch_rmniw$samples$Sigma[i, j, ],
         main = paste0("Sigma[", i, ",", j, "]"),
         xlab = "Federated rmniw Quantiles", ylab = "Single Server rmniw Quantiles",
         pch = 16, col = rgb(0.8, 0.4, 0.2, 0.4))
  abline(0, 1, col = "firebrick", lwd = 2)
}))
par(mfrow = c(1, 1))
```

---

# Execution Timing Benchmark

Comparing global sampling execution time across 10,000 posterior draws between custom R composition sampling and C++ accelerated sampling.

```{r timing_comparison}
n_bench <- 10000

t_custom <- system.time({
  res_bench_custom <- mniwSamplerFederated(client_data_list, m0, M0_inv, nu0, Psi0, n_samples = n_bench)
})["elapsed"]

t_rmniw <- system.time({
  res_bench_rmniw <- mniwSamplerFederated_rmniw(client_data_list, m0, M0_inv, nu0, Psi0, n_samples = n_bench)
})["elapsed"]

timing_df <- data.frame(
  Sampler_Implementation = c("Custom R Sampler (mniwSamplerFederated)", 
                             "C++ Accelerated Sampler (mniwSamplerFederated_rmniw)"),
  Elapsed_Time_Sec = c(as.numeric(t_custom), as.numeric(t_rmniw)),
  Speedup_Factor   = c(1.0, as.numeric(t_custom / t_rmniw))
)

kable(timing_df, digits = 4, caption = paste0("Federated Sampling Execution Time Comparison (", n_bench, " Samples)"))
```

The C++ sampling implementation via `mniw::rmniw` achieves substantial performance speedup while preserving exact statistical equivalence with the full centralized model.
