---
title: "Map Projections"
output: html_document
date: "2023-08-23"
---

```{r}
library(maps)
library(mapproj)
library(ggplot2)
library(sf)
```

1. Default world map
```{r}
p <- ggplot(map_data("world"),
            aes(long, lat, group = group)) +
  geom_polygon(fill = "white", colour = 1) +
  labs(title = "World Map")

p
```

2. Default CHINA and U.S. map
```{r}
world <- map_data("world")
china <- subset(world, region == "China")
usa <- map_data("usa")
mainland <- subset(usa, !(region %in% c("alaska", "hawaii")))


p_china <- ggplot(data = china, aes(long, lat, group = group)) +
  geom_polygon(fill = "white", colour = "black") +
  labs(title = "China Map")

p_mainland <- ggplot(data = mainland, aes(x = long, y = lat, group = group)) +
  geom_polygon(fill = "white", colour = "black") +
  labs(title = "USA Mainland Map") +
  theme(legend.position = "none")

p_china
p_mainland
```

3. "Mercator" Map Projection.\
**Advantage**: 1. Directional Accuracy. 2. Straight Rhumb Lines.\
**Disadvantage**: 1. Area Distortion. 2. Cannot Represent Polar Areas.
```{r}
p <- ggplot(map_data("world"),
            aes(long, lat, group = group)) +
      geom_polygon(fill = "white", colour = 1)

p + coord_map(xlim = c(-180, 180))
p + coord_sf() 
```

4."Orthographic" Map Projection\
The projection mimics the way distant celestial objects are viewed from a vast distance.
```{r}
p <- ggplot(map_data("world"),
            aes(long, lat, group = group)) +
      geom_polygon(fill = "white", colour = 1)

p + coord_map("orthographic") 
```

5."Fish Eye" Map Projection\
From the center point of the projection, distances are accurately represented. 
```{r}
p <- ggplot(map_data("world"),
            aes(long, lat, group = group)) +
      geom_polygon(fill = "white", colour = 1)

p + coord_map("fisheye", n = 4)
```

6."Sinusoidal" Map Projectionã€
Areas are represented accurately on the map.
```{r}
p <- ggplot(map_data("world"),
            aes(long, lat, group = group)) +
      geom_polygon(fill = "white", colour = 1)

p + coord_map("sinusoidal") 
```






