---
title: "Benchmarking Exact Bayesian Sampling Methods"
author: "Bayesian Computing Lab"
date: "`r Sys.Date()`"
output: 
  html_document:
    theme: cosmo
    toc: true
    toc_float: true
---

## 1. Introduction

In this laboratory, we compare the overall execution time of several analytical sampling wrappers for Bayesian Linear Regression. Because the external library wrappers (JAGS, NIMBLE, Stan) handle data packaging and model compilation internally, the timings for those frameworks reflect the full end-to-end processing cost.

## 2. Setup and Data Loading

We begin by loading the core libraries, setting a strict single-core limit to prevent memory overloads on barebones hardware, and loading the empirical dataset.

```{r setup, message=FALSE, warning=FALSE}
# Load required libraries
library(rjags)
library(nimble)
library(rstan)
library(ggplot2)

# Strictly limit Stan to 1 core to prevent Out-Of-Memory (OOM) compilation crashes
rstan_options(auto_write = TRUE)
options(mc.cores = 1)

# Load Data & Process Predictors
bench_dir <- getwd()
data_path <- normalizePath("../../data/data.txt", mustWork = FALSE)
rawData <- read.table(data_path, header = TRUE)

Y <- rawData$Y
X <- cbind(Intercept = 1, scale(as.matrix(rawData[, -1])))
N <- nrow(X)
P <- ncol(X)
n_samples <- 5000

# Shared Priors
mu0 <- rep(0, P)
V0_inv <- diag(1/100, P)
a0 <- 0.01
b0 <- 0.01

# Calculate QR Decomposition for the QR-optimized Raw Sampler
qr_decomp <- qr(X)
Q_mat <- qr.Q(qr_decomp)
R_mat <- qr.R(qr_decomp)
```

## 3. Execution Benchmarks

We source the raw R wrapper programs and time their full execution sequentially. We explicitly delete the heavy resulting objects and trigger aggressive garbage collection (`gc()`) between runs to keep the memory ceiling extremely low.

```{r benchmark, warning=FALSE, message=FALSE, results='hide'}
results_list <- list()

# 1. Base R Standard
source("../../raw/src/rawSampler.r")
time_raw <- system.time({
  res_raw <- rawSampler(y = Y, X = X, m0 = mu0, M0_inv = V0_inv, a0 = a0, b0 = b0, N = n_samples)
})
results_list[["Raw_Standard"]] <- time_raw["elapsed"]
rm(res_raw)
gc()

# 2. Base R QR-Optimized
source("../../raw/src/rawSamplerWithQR.r")
time_qr <- system.time({
  res_qr <- rawSamplerWithQR(y = Y, Q = Q_mat, R = R_mat, m0 = mu0, M0_inv = V0_inv, a0 = a0, b0 = b0, N = n_samples)
})
results_list[["Raw_QR"]] <- time_qr["elapsed"]
rm(res_qr)
gc()

# 3. JAGS Direct Wrapper
source("../../rjags/src/compositionSamplingJAGSDirect.r")
time_jags <- system.time({
  res_jags <- run_direct_jags_composition(X = X, Y = Y, n_samples = n_samples, a0 = a0, b0 = b0, mu0 = mu0, M0_inv = V0_inv)
})
results_list[["JAGS_Direct"]] <- time_jags["elapsed"]
rm(res_jags)
gc()

# 4. NIMBLE Direct Wrapper
source("../../nimble/src/compositionSamplingNimbleDirect.r")
time_nimble <- system.time({
  res_nimble <- run_direct_nimble_composition(X = X, Y = Y, n_samples = n_samples, a0 = a0, b0 = b0, mu0 = mu0, V0_inv = V0_inv)
})
results_list[["NIMBLE_Direct"]] <- time_nimble["elapsed"]
rm(res_nimble)
gc()

# 5. Stan Direct Wrapper
source("../../rstan/src/compositionSamplingStan.r")
time_stan <- system.time({
  res_stan <- run_direct_stan_composition(X = X, Y = Y, n_samples = n_samples, a0 = a0, b0 = b0, mu0 = mu0, V0_inv = V0_inv)
})
results_list[["Stan_Direct"]] <- time_stan["elapsed"]
rm(res_stan)
gc()
```

## 4. Benchmark Results

In the plot below, you will notice that NIMBLE and Stan exhibit significantly higher execution times. This is because our benchmarking dynamically invokes the wrappers exactly as written, which means `system.time()` is capturing both the **C++ translation** and the **system compilation overhead**. For a single execution script, this on-the-fly C++ compilation heavily dominates the actual statistical sampling time. Conversely, Base R executes natively in the interpreter, and JAGS utilizes pre-compiled C++ modules, allowing them to bypass this heavy compilation penalty entirely.

```{r display-benchmark, echo=FALSE, fig.width=10, fig.height=6}
# Combine results into a data frame
bench_df <- data.frame(
  Method = names(results_list),
  Time_Seconds = unlist(results_list)
)
rownames(bench_df) <- NULL
bench_df <- bench_df[order(bench_df$Time_Seconds), ]

# Print the numerical table
print(bench_df)

# Plot the results
ggplot(bench_df, aes(x = reorder(Method, Time_Seconds), y = Time_Seconds)) +
  geom_bar(stat = "identity", fill = "steelblue", color = "black", alpha = 0.8) +
  theme_minimal() +
  labs(
    title = "End-to-End Execution Time for Exact Bayesian Sampling Wrappers",
    subtitle = paste(n_samples, "Draws | Note: NIMBLE and Stan timings include C++ compilation overhead"),
    x = "Method",
    y = "Time (Seconds)"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 12))
```
