---
title: "Benchmarking Exact Bayesian Sampling Methods"
author: "Sudipto Banerjee"
date: "`r Sys.Date()`"
output: 
  html_document:
    theme: cosmo
    toc: true
    toc_float: true
---

## 1. Introduction

In this laboratory, we compare the computational performance of exact analytical sampling methods for Bayesian Linear Regression across several frameworks: Base R, JAGS, NIMBLE, and Stan. 

Instead of cluttering this report with the underlying algorithmic code, we have modularized our project. The sampler functions are sourced directly from our organized `src/` directories, enforcing clean software engineering practices.

## 2. Setup and Sourcing

First, we load our core libraries and source the refactored sampling scripts.

```{r setup, message=FALSE, warning=FALSE}
# Load required libraries
library(microbenchmark)
library(rjags)
library(nimble)
library(rstan)

# Configure Stan to utilize parallel cores
rstan_options(auto_write = TRUE)
options(mc.cores = parallel::detectCores())

# ---------------------------------------------------------
# Source Modular Samplers
# ---------------------------------------------------------
# 1. Raw Base R Samplers
source("../../raw/src/rawSampler.r")
source("../../raw/src/rawSamplerWithQR.r")

# 2. JAGS Samplers
source("../../rjags/src/compositionSamplingJAGSDirect.r")
source("../../rjags/src/compositionSamplingJAGS.r")

# 3. NIMBLE Samplers
source("../../nimble/src/compositionSamplingNimbleDirect.r")
source("../../nimble/src/compositionSamplingNimble.r")

# 4. Stan Samplers
source("../../rstan/src/compositionSamplingStan.r")
```

## 3. Data Loading

We load the empirical dataset for our benchmarking tests. We establish our target dimensions and the prior hyperparameters required by all of our functions.

```{r data-loading}
# Load the empirical dataset
# Assuming a standard tabular format (e.g., space or tab-separated)
dataset <- read.table("../../data/data.txt", header = TRUE)

# Parse Response (Y) and Predictors (X)
# NOTE: This assumes Y is the FIRST column. Change to dataset[, ncol(dataset)] if Y is last.
Y <- as.numeric(dataset[, 1])
X <- as.matrix(dataset[, -1])

# Extract Dimensions
N <- nrow(X)
P <- ncol(X)

# Define Shared Prior Parameters
m0 <- rep(0, P)
M0_inv <- diag(0.01, P)
a0 <- 0.01
b0 <- 0.01

# Calculate QR Decomposition for the QR-optimized Raw Sampler
qr_decomp <- qr(X)
Q_mat <- qr.Q(qr_decomp)
R_mat <- qr.R(qr_decomp)

# Specify number of posterior draws requested
n_draws <- 5000
```

## 4. Benchmarking Execution Speeds

We now utilize the `microbenchmark` package to compare the execution times of our different sampling strategies. 

**Note on Compilation:** Functions utilizing external C++ compilers (Stan and NIMBLE) will incur a one-time compilation overhead during the benchmark. The raw R functions rely on optimized underlying BLAS/LAPACK Fortran routines.

```{r benchmark, warning=FALSE, message=FALSE, cache=FALSE, results='hide'}
# Run the benchmark (times = 5 to account for compilation overhead in C++ methods)
benchmark_results <- microbenchmark(
  
  # Base R Implementations
  Raw_Standard = rawSampler(
    y = Y, X = X, m0 = m0, M0_inv = M0_inv, a0 = a0, b0 = b0, N = n_draws
  ),
  
  Raw_QR = rawSamplerWithQR(
    y = Y, Q = Q_mat, R = R_mat, m0 = m0, M0_inv = M0_inv, a0 = a0, b0 = b0, N = n_draws
  ),
  
  # JAGS Implementations
  JAGS_Direct = run_direct_jags_composition(
    X = X, Y = Y, n_samples = n_draws, a0 = a0, b0 = b0, m0 = m0, M0_inv = M0_inv
  ),
  
  JAGS_Replicated = run_replicated_jags_composition(
    X = X, Y = Y, M_samples = n_draws, a0 = a0, b0 = b0, m0 = m0, M0_inv = M0_inv
  ),
  
  # NIMBLE Implementations
  NIMBLE_Direct = run_direct_nimble_composition(
    X = X, Y = Y, n_samples = n_draws, a0 = a0, b0 = b0, m0 = m0, M0_inv = M0_inv
  ),
  
  NIMBLE_Replicated = run_replicated_nimble_composition(
    X = X, Y = Y, M_samples = n_draws, a0 = a0, b0 = b0, m0 = m0, M0_inv = M0_inv
  ),
  
  # Stan Implementation
  Stan_Direct = run_direct_stan_composition(
    X = X, Y = Y, n_samples = n_draws, a0 = a0, b0 = b0, m0 = m0, M0_inv = M0_inv
  ),
  
  times = 5
)
```

### Benchmark Results

```{r display-benchmark, echo=FALSE}
# Print formatted results
print(benchmark_results, unit = "s", signif = 4)
```

```{r plot-benchmark, echo=FALSE, fig.width=10, fig.height=6}
# Visualize the benchmark distributions
library(ggplot2)
autoplot(benchmark_results) + 
  theme_minimal() +
  labs(
    title = "Execution Time Comparison for Exact Bayesian Sampling",
    subtitle = paste(n_draws, "Posterior Draws |", N, "Observations |", P, "Predictors"),
    x = "Method",
    y = "Time (Seconds)"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 12))
```
