---
title: "Exploratory data analysis for North Carolina Sudden Infant Death Syndrome (SIDS)"
output: html_document
---

```{r global_options, echo=FALSE}
knitr::opts_chunk$set(comment = NA, tidy = TRUE)
```

First we clear memory. This is always a good practice to avoid possible conflicts with objects in previously saved workspaces if any.
```{r, clean_memory}
rm(list=ls())
```

Install a few packages that we will use.
```{r, install_packages}
library(spdep)
library(maptools)
library(RColorBrewer)
library(spatialreg)
```

Load the data from a shapefile in the "spdep" package.
```{r, load_data}
nc.sids <- st_read(system.file("shapes/sids.shp", package="spData")[1], quiet=TRUE) 
st_crs(nc.sids) = "+proj=longlat +ellps=clrk66"
row.names(nc.sids) <- as.character(nc.sids$FIPSNO)
```

Extract some attributes of the spatial data.
```{r, attributes}
nc.county.id = attr(ncCC89.nb, "region.id")
nc.no.neighbors = card(ncCC89.nb)
nc.islands = as.character(nc.sids[card(ncCC89.nb) == 0, ]$NAME)
```

Create a map of raw SIDS rates. First create a vector of rates from counts.
```{r, map}
nc.sids.rates.raw = probmap(nc.sids$SID79, nc.sids$BIR79)$raw
##Alternatvely, one may want to use Freeman-Tukey transformations
nc.sids.rates.FT = sqrt(1000) * (sqrt(nc.sids$SID79/nc.sids$BIR79) + sqrt((nc.sids$SID79 + 1)/nc.sids$BIR79))
##Append the rates we want to plot to the nc.sids data frame
nc.sids$rates.FT = nc.sids.rates.FT 
brks = c(0, 2.0, 3.0, 3.5, 6.0)
```

Compute Moran's I and Geary's C
```{r, Morans I and Gearys C}
ncCC89.listw = nb2listw(ncCC89.nb, style="B", zero.policy=TRUE)
##ncCR85.listw = nb2listw(ncCR85.nb, style="W", zero.policy=TRUE)
nc.sids.moran.out = moran.test(nc.sids.rates.FT, listw=ncCC89.listw, zero.policy=TRUE) 
nc.sids.geary.out = geary.test(nc.sids.rates.FT, listw=ncCC89.listw, zero.policy=TRUE) 
```

SAR model regressing rates.FT on NWBIR79.FT. Begin with a Freeman-Tukey transformation on non-white birth rates
```{r, SAR model}
nc.sids.nwbir.FT = sqrt(1000) * (sqrt(nc.sids$NWBIR79/nc.sids$BIR79) + sqrt((nc.sids$NWBIR79 + 1)/nc.sids$BIR79))
nc.sids$nwbir.FT = nc.sids.nwbir.FT
## Use errorsarlm and lagsarlm
##nc.sids.sar.out = errorsarlm(rates.FT~ nwbir.FT, data=nc.sids, listw=ncCC89.listw, zero.policy=TRUE)
nc.sids.sar.out = spautolm(rates.FT~ nwbir.FT, data=nc.sids, family="SAR", listw=ncCC89.listw, zero.policy=TRUE)
nc.sids.lagsar.out = lagsarlm(rates.FT~ nwbir.FT, data=nc.sids, listw=ncCC89.listw, zero.policy=TRUE)
nc.sids.sar.fitted = fitted(nc.sids.sar.out)
nc.sids$fitted.sar = nc.sids.sar.fitted
```

CAR model regressing rates.FT on NWBIR79.FT
```{r, CAR model}
nc.sids.car.out = spautolm(rates.FT~ nwbir.FT, data=nc.sids, family="CAR", listw=ncCC89.listw, zero.policy=TRUE)
nc.sids.car.fitted = fitted(nc.sids.car.out)
nc.sids$fitted.car = nc.sids.car.fitted
```

Draw the maps using the maps function
```{r, draw_maps}
library(maps)
library(classInt)

color.pallete = rev(brewer.pal(4,"RdBu"))
class.raw = classIntervals(var=nc.sids$rates.FT, n=4, style="fixed", fixedBreaks=brks, dataPrecision=4)
color.code.raw = findColours(class.raw, color.pallete)
class.fitted = classIntervals(var=nc.sids$fitted.sar, n=4, style="fixed", fixedBreaks=brks, dataPrecision=4)
color.code.fitted = findColours(class.fitted, color.pallete)

leg.txt = c("<2.0", "2.0-3.0", "3.0-3.5",">3.5")

#pdf(file="SidsSar.pdf")
par(mfrow=c(2,1), oma = c(0,0,4,0) + 0.1, mar = c(0,0,1,0) + 0.1)
plot(nc.sids[c("geometry","rates.FT")], col=color.code.raw, main = "", key.pos = NULL, reset = FALSE)
title("a) Raw Freeman-Tukey transformed SIDS rates" )
legend("bottomleft", legend=leg.txt, cex=1.25, bty="n", horiz = FALSE, fill = color.pallete)
plot(nc.sids[c("geometry","fitted.sar")], col=color.code.fitted, main = "", key.pos = NULL, reset = FALSE)
title("b) Fitted SIDS rates from SAR model")
#dev.off()
```

