---
title: "Bayesian Linear Regression via QR Decomposition"
subtitle: "Theory, Composition Sampling Algorithms, and Raw R Implementation"
author: "Sudipto Banerjee"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Introduction and Mathematical Reparameterization

In standard linear regression modeling, parameter estimation and sampling can suffer from severe numerical instability when predictors operate on drastically different scales or display strong multicollinearity. This instability arises primarily when evaluating matrix products such as $X^\top X$ or inverting ill-conditioned covariance matrices.

To achieve optimal numerical stability without altering the physical units or requiring ad-hoc predictor standardization, we apply a **thin QR decomposition** to the design matrix $X \in \mathbb{R}^{n \times p}$:

$$X = QR$$

where:

* $Q \in \mathbb{R}^{n \times p}$ is an orthogonal matrix satisfying $Q^\top Q = I_p$.
* $R \in \mathbb{R}^{p \times p}$ is an upper-triangular matrix with positive diagonal entries.

## Model Reparameterization

Substituting $X = QR$ into the linear regression model $y = X\beta + \varepsilon$ yields:

$$y = Q(R\beta) + \varepsilon = Q\theta + \varepsilon, \quad \varepsilon \sim \mathcal{N}_n(0_{n \times 1}, \, \sigma^2 I_n)$$

where the transformed parameter vector $\theta \in \mathbb{R}^{p \times 1}$ is defined as:

$$\theta = R\beta \iff \beta = R^{-1}\theta$$

Because $Q^\top Q = I_p$, the condition number of the design matrix $Q$ in $\theta$-space is identically $1$. This orthogonality completely decouples the numerical updates during posterior inference.

---

# Efficient Computation of Prior Precision ($M_\theta^{-1}$)

Under a conjugate Gaussian prior on the original parameters $\beta \mid \sigma^2 \sim \mathcal{N}_p\left(m_0, \, \sigma^2 M_0^{-1}\right)$, the induced prior on $\theta = R\beta$ is:

$$\theta \mid \sigma^2 \sim \mathcal{N}_p\left(m_\theta, \, \sigma^2 M_\theta^{-1}\right)$$

where:

$$m_\theta = R m_0, \quad M_\theta = R M_0 R^\top$$

The joint posterior update requires the prior precision factor $M_\theta^{-1} = (R^\top)^{-1} M_0^{-1} R^{-1}$. Explicitly inverting triangular matrices introduces unnecessary numerical loss. Instead, we evaluate $M_\theta^{-1}$ using **two sequential lower-triangular matrix solves** with $R^\top$:

1. **First Solve:** Solve the lower-triangular system $R^\top A = M_0^{-1}$ for $A \in \mathbb{R}^{p \times p}$ via forward-substitution:
   $$A = (R^\top)^{-1} M_0^{-1} \quad \text{(R code: `A <- forwardsolve(t(R), M0_inv)`)}$$

2. **Second Solve:** Note that $M_\theta^{-1} = A R^{-1} \implies M_\theta^{-1} R = A$. Taking the transpose yields:
   $$R^\top M_\theta^{-1} = A^\top \quad \text{(R code: `M_theta_inv <- forwardsolve(t(R), t(A))`)}$$

This approach completely eliminates explicit matrix inversions, preserves numerical precision, and operates in $O(p^3)$ triangular solves.

---

# Composition Sampling Algorithms for Parameter Inference

## Algorithm 1: General Conjugate Prior Case (User Supplies $M_0^{-1}$)

**Inputs:** Response vector $y \in \mathbb{R}^n$, unscaled design matrix $X \in \mathbb{R}^{n \times p}$, prior precision factor $M_0^{-1} \in \mathbb{R}^{p \times p}$, prior location $m_0 \in \mathbb{R}^{p \times 1}$, Inverse-Gamma hyperparameters $a_0, b_0$, and requested draw count $N$.

1. Perform QR decomposition on $X$ to obtain $Q \in \mathbb{R}^{n \times p}$ and upper-triangular $R \in \mathbb{R}^{p \times p}$.
2. Solve $R^\top m_\theta = m_0$ for $m_\theta$ using lower-triangular forward-substitution.
3. Compute $M_\theta^{-1} = (R^\top)^{-1} M_0^{-1} R^{-1}$ efficiently **without matrix inversion** using two triangular solves:
   a. Solve $R^\top A = M_0^{-1}$ for $A \in \mathbb{R}^{p \times p}$ via forward-substitution.
   b. Solve $R^\top M_\theta^{-1} = A^\top$ for $M_\theta^{-1}$ via forward-substitution.
4. Form posterior precision factor $M_{\theta, n}^{-1} = M_\theta^{-1} + I_p$, and compute its upper-triangular Cholesky factor $R_{\theta, n}$:
   $$R_{\theta, n}^\top R_{\theta, n} = M_{\theta, n}^{-1}$$
5. Compute location vector $m_{\theta, n} = m_\theta + Q^\top y$. Obtain posterior mean vector $\mu_{\theta, n} = M_{\theta, n} m_{\theta, n}$ without matrix inversion via two triangular solves:
   $$\text{Solve } R_{\theta, n}^\top w = m_{\theta, n} \text{ for } w, \quad \text{then solve } R_{\theta, n} \mu_{\theta, n} = w \text{ for } \mu_{\theta, n}$$
6. Update Inverse-Gamma shape and scale parameters:
   $$a_n = a_0 + \frac{n}{2}, \quad b_n = b_0 + \frac{1}{2}\left[ m_0^\top M_0 m_0 + y^\top y - w^\top w \right]$$
7. Vectorized sampling across all $N$ draws:
   a. Draw $\sigma^{2(1:N)} \sim \text{Inv-Gamma}\left(a_n, \, b_n\right)$.
   b. Draw matrix $Z \in \mathbb{R}^{p \times N}$ from standard Normal and solve $R_{\theta, n} V = Z$ via back-substitution to form $\Theta \in \mathbb{R}^{p \times N}$:
      $$\Theta = \mu_{\theta, n} \mathbf{1}_N^\top + V \cdot \text{diag}\left(\sigma^{(1:N)}\right)$$
   c. Recover natural regression parameters $\mathcal{B} \in \mathbb{R}^{N \times p}$ by solving $R \mathcal{B}^\top = \Theta$ via triangular back-substitution:
      $$\mathcal{B} = \text{backsolve}(R, \Theta)^\top$$
8. Return posterior samples $\left(\mathcal{B}, \sigma^{2(1:N)}\right)$.

---

## Algorithm 2: Non-Informative (Jeffreys) Flat Prior Case

**Inputs:** Response vector $y \in \mathbb{R}^n$, unscaled design matrix $X \in \mathbb{R}^{n \times p}$, and requested draw count $N$.

1. Perform QR decomposition on $X$ to obtain $Q \in \mathbb{R}^{n \times p}$ and upper-triangular $R \in \mathbb{R}^{p \times p}$.
2. Compute orthogonal OLS estimate and sum of squared errors:
   $$\hat{\theta}_{\text{OLS}} = Q^\top y, \quad \text{SSE} = y^\top y - \hat{\theta}_{\text{OLS}}^\top \hat{\theta}_{\text{OLS}}$$
3. Set posterior Inverse-Gamma parameters:
   $$a_n = \frac{n - p}{2}, \quad b_n = \frac{\text{SSE}}{2}$$
4. Vectorized sampling across all $N$ draws:
   a. Draw $\sigma^{2(1:N)} \sim \text{Inv-Gamma}\left(a_n, \, b_n\right)$.
   b. Draw $\Theta = \hat{\theta}_{\text{OLS}} \mathbf{1}_N^\top + Z \cdot \text{diag}\left(\sigma^{(1:N)}\right)$, where $Z \in \mathbb{R}^{p \times N} \sim \mathcal{N}(0, 1)$.
   c. Recover natural regression parameters $\mathcal{B} = \text{backsolve}(R, \Theta)^\top$.
5. Return posterior samples $\left(\mathcal{B}, \sigma^{2(1:N)}\right)$.

---

# Posterior Predictive Sampling for New Data ($X_{\text{new}}$)

Once posterior samples $\left(\mathcal{B}, \sigma^{2(1:N)}\right)$ are generated, we can make predictions for new covariate observations $X_{\text{new}} \in \mathbb{R}^{n_{\text{new}} \times p}$.

## Conceptual Note: Why We Do Not Decompose $X_{\text{new}}$

It is a common question whether one should perform a separate QR decomposition $X_{\text{new}} = Q_{\text{new}} R_{\text{new}}$. The answer is **no**:

1. **Numerical Stability:** QR decomposition is applied to $X$ during model training to prevent numerical breakdown in matrix inversions (e.g., inverting $X^\top X$). Prediction is a forward matrix-vector multiplication: $\mu_{\text{new}}^{(s)} = X_{\text{new}}\beta^{(s)}$. This operation requires zero matrix solves or inversions and is inherently stable regardless of predictor scaling in $X_{\text{new}}$.
2. **Mathematical Incompatibility:** The reparameterized vector $\theta$ is tied to the training matrix factor via $R_{\text{train}}\beta = \theta$. A separate decomposition $X_{\text{new}} = Q_{\text{new}} R_{\text{new}}$ yields a completely different factor $R_{\text{new}} \neq R_{\text{train}}$. Consequently, $Q_{\text{new}}\theta^{(s)} \neq X_{\text{new}}\beta^{(s)}$, producing mathematically invalid predictions.

---

# Clarification: Centered Intercept ($\alpha$) vs. Raw Intercept ($\beta_0$)

When fitting linear models with mean-centered predictors $X_c = X - \bar{X}$, it is crucial to distinguish between two intercept quantities:

1. **Centered / Orthogonal Intercept ($\alpha$):**
   The intercept parameter for centered predictors represents the expected response at the average predictor values ($\bar{X}$). Because the centered covariates have sample mean zero, $\alpha$ anchors directly around the sample mean of the response variable:
   $$\alpha = \bar{y} \approx 11.00 \quad \text{(for empirical data)}$$
2. **Raw Uncentered Intercept ($\beta_0$):**
   The intercept transformed back to raw uncentered coordinates ($X = 0$). When raw covariates operate on large baseline scales (such as geographic coordinates like Latitude $\approx 35^\circ$ or Longitude $\approx -80^\circ$), extrapolating back to $(0,0)$ yields:
   $$\beta_0 = \alpha - \sum_{j=2}^p \bar{X}_j \beta_j \approx -500$$

Both quantities are mathematically consistent; $\alpha \approx 11$ measures local response height at average covariate values, whereas $\beta_0 \approx -500$ measures the response plane evaluated at raw $X = 0$.

---

# Raw R Code for Bayesian Inference with QR Decomposition

In this section, we evaluate the QR-based sampling workflow using executable R code. We compare the computational accuracy of `rawSamplerWithQR()` against `rawSampler.r`.

## Setup and Source Files

We source the required sampler routines directly from the `../src/` directory:

```{r setup-libs, message=FALSE, warning=FALSE}
# Source raw sampler implementations from ../src/
source("../src/rawSampler.r")
source("../src/rawSamplerWithQR.r")
source("../src/rawPosteriorPredictive.r")

library(knitr)
```

To test numerical stability, we simulate a dataset where the predictor variables operate on vastly different scales ($X_1 \in [0, 1]$ versus $X_2 \in [10^4, 10^5]$).

```{r simulate-data}
set.seed(2026)
n <- 250
p <- 3
N_samples <- 10000

# Predictors with extreme scale differences
X_raw <- cbind(
  Intercept = 1,
  X1 = runif(n, 0, 1),
  X2 = runif(n, 1e4, 1e5)
)

true_beta <- c(12.5, -4.2, 0.00085)
true_sigma2 <- 4.0

# Generate response y
y <- as.numeric(X_raw %*% true_beta + rnorm(n, mean = 0, sd = sqrt(true_sigma2)))
```

---

## Posterior Inference and Comparisons

We fit both models under a non-informative flat Jeffreys prior ($M_0^{-1} = 0_{p \times p}$, $m_0 = 0_{p \times 1}$, $a_0 = -p/2$, $b_0 = 0$):

1. **Standardized Predictor Sampler (`rawSampler.r`):** Centers and scales $X$ prior to sampling, then converts back to original units.
2. **QR Reparameterized Sampler (`rawSamplerWithQR.r`):** Operates on centered predictors using QR decomposition ($X_c = QR$).

```{r fit-models}
# Set flat prior parameters
m0 <- rep(0, p)
M0_inv <- matrix(0, p, p)
a0 <- -p / 2
b0 <- 0

# -----------------------------------------------------------------------------
# Method 1: Standardized Sampler (rawSampler.r)
# -----------------------------------------------------------------------------
X_center <- colMeans(X_raw[, -1])
X_scale  <- apply(X_raw[, -1], 2, sd)
X_scaled <- cbind(Intercept = 1, scale(X_raw[, -1]))

res_scaled <- rawSampler(y, X_scaled, m0, M0_inv, a0, b0, N_samples)

# Centered intercept draws (alpha ~ y_bar)
alpha_scaled_draws <- res_scaled$beta[, 1]

# Unscale Beta draws back to raw data units (Vectorized)
beta_scaled_draws <- res_scaled$beta
beta_unscaled_draws <- beta_scaled_draws
beta_unscaled_draws[, 2:3] <- sweep(beta_scaled_draws[, 2:3], 2, X_scale, "/")
beta_unscaled_draws[, 1]   <- beta_scaled_draws[, 1] - 
  rowSums(sweep(beta_scaled_draws[, 2:3], 2, X_center / X_scale, "*"))

# -----------------------------------------------------------------------------
# Method 2: QR Sampler on Centered Predictors (rawSamplerWithQR.r)
# -----------------------------------------------------------------------------
X_centered <- cbind(Intercept = 1, scale(X_raw[, -1], center = TRUE, scale = FALSE))
qr_X <- qr(X_centered)
Q <- qr.Q(qr_X)
R <- qr.R(qr_X)

res_qr <- rawSamplerWithQR(y, Q, R, m0, M0_inv, a0, b0, N_samples)

# Centered intercept draws (alpha ~ y_bar)
alpha_qr_draws <- res_qr$beta[, 1]

# Convert centered QR Beta draws to raw unscaled predictors (Vectorized)
beta_qr_draws <- res_qr$beta
beta_unscaled_qr <- beta_qr_draws
beta_unscaled_qr[, 1] <- beta_qr_draws[, 1] - rowSums(sweep(beta_qr_draws[, 2:3], 2, X_center, "*"))
```

### Tabular Posterior Summary Comparison

Below is a comparison of posterior means, standard deviations, and 95% credible intervals across both methods, displaying both the centered intercept ($\alpha$) and the uncentered raw intercept ($\beta_0$):

```{r posterior-table, echo=FALSE}
calc_stats <- function(draws_matrix, sigma2_vec) {
  means <- colMeans(draws_matrix)
  sds   <- apply(draws_matrix, 2, sd)
  q025  <- apply(draws_matrix, 2, quantile, probs = 0.025)
  q975  <- apply(draws_matrix, 2, quantile, probs = 0.975)
  
  data.frame(
    Mean = sprintf("%.6f", means),
    SD   = sprintf("%.6f", sds),
    `2.5%` = sprintf("%.6f", q025),
    `97.5%` = sprintf("%.6f", q975),
    check.names = FALSE
  )
}

draws_scaled_full <- cbind(Centered_Alpha = alpha_scaled_draws, beta_unscaled_draws, sigma2 = res_scaled$sigma2)
draws_qr_full     <- cbind(Centered_Alpha = alpha_qr_draws, beta_unscaled_qr, sigma2 = res_qr$sigma2)

stats_scaled <- calc_stats(draws_scaled_full, res_scaled$sigma2)
stats_qr     <- calc_stats(draws_qr_full, res_qr$sigma2)

param_names <- c("alpha (Centered Intercept ~ y_bar)", "beta_0 (Raw Intercept at X=0)", "beta_1 (X1)", "beta_2 (X2)", "sigma^2")

comp_table <- data.frame(
  Parameter = rep(param_names, each = 2),
  Method    = rep(c("rawSampler (Standardized X)", "rawSamplerWithQR (Centered X)"), times = length(param_names)),
  Mean      = c(rbind(stats_scaled$Mean, stats_qr$Mean)),
  SD        = c(rbind(stats_scaled$SD, stats_qr$SD)),
  `2.5%`    = c(rbind(stats_scaled$`2.5%`, stats_qr$`2.5%`)),
  `97.5%`   = c(rbind(stats_scaled$`97.5%`, stats_qr$`97.5%`)),
  check.names = FALSE
)

kable(comp_table, caption = "Posterior Estimation Comparison: rawSampler vs. rawSamplerWithQR", align = "lccccc")
```

### Quantile-Quantile (Q-Q) Plots for Posterior Parameter Draws

To visually verify that both samplers produce identical empirical posterior distributions, we construct Q-Q plots for each parameter draw:

```{r posterior-qqplots, fig.width=9, fig.height=8, fig.align='center'}
par(mfrow = c(2, 2), mar = c(4.5, 4.5, 2.5, 1))

qqplot(alpha_scaled_draws, alpha_qr_draws,
       xlab = "rawSampler Centered Intercept", ylab = "rawSamplerWithQR Centered Intercept",
       main = expression("Q-Q Plot for Centered Intercept (" * alpha * " ~ " * bar(y) * ")"))
abline(0, 1, col = "firebrick", lwd = 2)

qqplot(beta_unscaled_draws[, 1], beta_unscaled_qr[, 1],
       xlab = "rawSampler Raw Intercept", ylab = "rawSamplerWithQR Raw Intercept",
       main = expression("Q-Q Plot for Raw Intercept (" * beta[0] * ")"))
abline(0, 1, col = "firebrick", lwd = 2)

qqplot(beta_unscaled_draws[, 2], beta_unscaled_qr[, 2],
       xlab = "rawSampler Draw Quantiles", ylab = "rawSamplerWithQR Draw Quantiles",
       main = expression("Q-Q Plot for " * beta[1]))
abline(0, 1, col = "firebrick", lwd = 2)

qqplot(res_scaled$sigma2, res_qr$sigma2,
       xlab = "rawSampler Draw Quantiles", ylab = "rawSamplerWithQR Draw Quantiles",
       main = expression("Q-Q Plot for " * sigma^2))
abline(0, 1, col = "firebrick", lwd = 2)
```

The sample quantiles lie directly along the $45^\circ$ reference line ($y = x$), demonstrating exact empirical distribution equivalence between `rawSampler()` and `rawSamplerWithQR()`.

---

## Posterior Predictive Inference and Comparisons

We evaluate posterior predictive sampling for a new matrix of unscaled test observations $X_{\text{new}} \in \mathbb{R}^{m_{\text{test}} \times p}$ using `rawPosteriorPredictive()`:

```{r predictive-sampling}
m_test <- 100
X_new_raw <- cbind(
  Intercept = 1,
  X1 = runif(m_test, 0, 1),
  X2 = runif(m_test, 1e4, 1e5)
)

# 1. Predictions using rawSamplerWithQR Draws
Y_pred_qr <- rawPosteriorPredictive(X_new_raw, beta_unscaled_qr, res_qr$sigma2)

# 2. Predictions using rawSampler Draws (unscaled parameters)
Y_pred_scaled <- rawPosteriorPredictive(X_new_raw, beta_unscaled_draws, res_scaled$sigma2)
```

### Posterior Predictive Quantile-Quantile (Q-Q) Plots

We select four individual test observations across $X_{\text{new}}$ and compare their posterior predictive distributions $y_{\text{new}}$:

```{r predictive-qqplots, fig.width=9, fig.height=8, fig.align='center'}
par(mfrow = c(2, 2), mar = c(4.5, 4.5, 2.5, 1))

test_indices <- c(1, 25, 50, 100)
invisible(lapply(test_indices, function(idx) {
  qqplot(Y_pred_scaled[idx, ], Y_pred_qr[idx, ],
         xlab = "rawSampler Predictive Quantiles",
         ylab = "rawSamplerWithQR Predictive Quantiles",
         main = paste("Posterior Predictive Q-Q Plot: Obs #", idx))
  abline(0, 1, col = "dodgerblue3", lwd = 2)
}))
```

---

## Empirical Application on Full Dataset (`data.txt`)

To validate both sampler implementations on real-world empirical data, we analyze the full dataset (`data.txt`) containing $N = 1,369$ observations across $p = 9$ parameters (intercept, six spatial/structural predictors `X1`–`X6`, plus `Lat` and `Longi`).

We fit both `rawSampler.r` and `rawSamplerWithQR.r` under a non-informative Jeffreys flat prior. By mean-centering the predictor covariates prior to QR decomposition, the centered intercept parameter $\alpha$ anchors cleanly around the global response mean ($\bar{y} \approx 11.00$), whereas the raw intercept $\beta_0$ equals $\approx -500$ due to extrapolating spatial predictors back to zero coordinates.

```{r empirical-analysis}
# Read full empirical dataset
data_path <- if (file.exists("../../data/data.txt")) "../../data/data.txt" else "data.txt"
full_data <- read.table(data_path, header = TRUE)

y_emp <- full_data[[1]]
X_emp_covariates <- as.matrix(full_data[, -1])

n_emp <- length(y_emp)
p_emp <- ncol(X_emp_covariates) + 1

X_center_emp <- colMeans(X_emp_covariates)
X_scale_emp  <- apply(X_emp_covariates, 2, sd)

# Flat Jeffreys prior setup
m0_emp <- rep(0, p_emp)
M0_inv_emp <- matrix(0, p_emp, p_emp)
a0_emp <- -p_emp / 2
b0_emp <- 0

# -----------------------------------------------------------------------------
# Method 1: Standardized Sampler on Empirical Data (rawSampler.r)
# -----------------------------------------------------------------------------
X_scaled_emp <- cbind(Intercept = 1, scale(X_emp_covariates))

res_scaled_emp <- rawSampler(y_emp, X_scaled_emp, m0_emp, M0_inv_emp, a0_emp, b0_emp, N_samples)

# Centered intercept draws (alpha ~ y_bar ~ 11.00)
alpha_scaled_emp <- res_scaled_emp$beta[, 1]

# Unscale Beta draws back to natural physical units (Vectorized)
beta_scaled_draws_emp <- res_scaled_emp$beta
beta_unscaled_draws_emp <- beta_scaled_draws_emp
beta_unscaled_draws_emp[, -1] <- sweep(beta_scaled_draws_emp[, -1], 2, X_scale_emp, "/")
beta_unscaled_draws_emp[, 1]  <- beta_scaled_draws_emp[, 1] - 
  rowSums(sweep(beta_scaled_draws_emp[, -1], 2, X_center_emp / X_scale_emp, "*"))

# -----------------------------------------------------------------------------
# Method 2: QR Sampler on Centered Empirical Predictors (rawSamplerWithQR.r)
# Mean-centering predictors anchors centered intercept alpha at y_bar (~11.00)
# -----------------------------------------------------------------------------
X_centered_emp <- cbind(Intercept = 1, scale(X_emp_covariates, center = TRUE, scale = FALSE))
qr_X_emp <- qr(X_centered_emp)
Q_emp <- qr.Q(qr_X_emp)
R_emp <- qr.R(qr_X_emp)

res_qr_emp <- rawSamplerWithQR(y_emp, Q_emp, R_emp, m0_emp, M0_inv_emp, a0_emp, b0_emp, N_samples)

# Centered intercept draws (alpha ~ y_bar ~ 11.00)
alpha_qr_emp <- res_qr_emp$beta[, 1]

# Convert centered QR Beta draws to raw unscaled predictors (Vectorized)
beta_qr_draws_emp <- res_qr_emp$beta
beta_unscaled_qr_emp <- beta_qr_draws_emp
beta_unscaled_qr_emp[, 1] <- beta_qr_draws_emp[, 1] - 
  rowSums(sweep(beta_qr_draws_emp[, -1], 2, X_center_emp, "*"))
```

### Posterior Summaries for Empirical Dataset

Below is the comparative posterior inference table for all parameters, showing both the centered intercept ($\alpha \approx 11.00$) and the raw uncentered intercept ($\beta_0 \approx -500$):

```{r empirical-table, echo=FALSE}
draws_scaled_emp_full <- cbind(Centered_Alpha = alpha_scaled_emp, beta_unscaled_draws_emp, sigma2 = res_scaled_emp$sigma2)
draws_qr_emp_full     <- cbind(Centered_Alpha = alpha_qr_emp, beta_unscaled_qr_emp, sigma2 = res_qr_emp$sigma2)

stats_scaled_emp <- calc_stats(draws_scaled_emp_full, res_scaled_emp$sigma2)
stats_qr_emp     <- calc_stats(draws_qr_emp_full, res_qr_emp$sigma2)

param_names_emp <- c("alpha (Centered Intercept ~ y_bar)", "beta_0 (Raw Intercept at X=0)", colnames(X_emp_covariates), "sigma^2")

comp_table_emp <- data.frame(
  Parameter = rep(param_names_emp, each = 2),
  Method    = rep(c("rawSampler (Standardized X)", "rawSamplerWithQR (Centered X)"), times = length(param_names_emp)),
  Mean      = c(rbind(stats_scaled_emp$Mean, stats_qr_emp$Mean)),
  SD        = c(rbind(stats_scaled_emp$SD, stats_qr_emp$SD)),
  `2.5%`    = c(rbind(stats_scaled_emp$`2.5%`, stats_qr_emp$`2.5%`)),
  `97.5%`   = c(rbind(stats_scaled_emp$`97.5%`, stats_qr_emp$`97.5%`)),
  check.names = FALSE
)

kable(comp_table_emp, caption = "Empirical Posterior Estimation Comparison: rawSampler vs. rawSamplerWithQR (data.txt)", align = "lccccc")
```

### Quantile-Quantile (Q-Q) Plots for Empirical Parameter Draws

To verify distribution-level equivalence across all parameters on the full dataset, we generate Q-Q plots for each coefficient draw:

```{r empirical-qqplots, fig.width=10, fig.height=9, fig.align='center'}
par(mfrow = c(3, 4), mar = c(4, 4, 2.5, 1))

# Centered Intercept Alpha
qqplot(alpha_scaled_emp, alpha_qr_emp,
       xlab = "rawSampler Draw Quantiles", 
       ylab = "rawSamplerWithQR Draw Quantiles",
       main = expression("Q-Q Plot: " * alpha * " (~ " * bar(y) * ")"))
abline(0, 1, col = "firebrick", lwd = 2)

# Raw Intercept Beta_0 and Slopes
invisible(lapply(1:p_emp, function(i) {
  p_name <- if (i == 1) "beta_0 (Raw Intercept)" else colnames(X_emp_covariates)[i - 1]
  qqplot(beta_unscaled_draws_emp[, i], beta_unscaled_qr_emp[, i],
         xlab = "rawSampler Draw Quantiles", 
         ylab = "rawSamplerWithQR Draw Quantiles",
         main = paste("Q-Q Plot:", p_name))
  abline(0, 1, col = "firebrick", lwd = 2)
}))

# Sigma^2
qqplot(res_scaled_emp$sigma2, res_qr_emp$sigma2,
       xlab = "rawSampler Draw Quantiles", 
       ylab = "rawSamplerWithQR Draw Quantiles",
       main = expression("Q-Q Plot for " * sigma^2))
abline(0, 1, col = "firebrick", lwd = 2)
```

The sample quantiles strictly adhere to the identity diagonal ($y = x$) across all parameters on the empirical dataset, confirming that `rawSamplerWithQR()` aligns perfectly with `rawSampler()`.

---

### Empirical Posterior Predictive Sampling and Diagnostic Checks

To complete the predictive inference for the empirical application (`data.txt`), we generate posterior predictive draws $Y_{\text{pred}} \in \mathbb{R}^{n_{\text{emp}} \times N}$ across all $n_{\text{emp}} = 1,369$ empirical observations using `rawPosteriorPredictive()`.

As established conceptually, we pass the raw uncentered design matrix $X_{\text{emp, raw}} = [1 \quad X_{\text{covariates}}]$ together with the unscaled posterior parameter samples $(\mathcal{B}, \sigma^{2(1:N)})$.

```{r empirical-predictive-sampling}
# Construct raw unscaled design matrix for empirical observations
X_emp_raw <- cbind(Intercept = 1, X_emp_covariates)

# 1. Generate posterior predictive distributions using QR Sampler draws
Y_pred_qr_emp <- rawPosteriorPredictive(X_emp_raw, beta_unscaled_qr_emp, res_qr_emp$sigma2)

# 2. Generate posterior predictive distributions using Standard Sampler draws
Y_pred_scaled_emp <- rawPosteriorPredictive(X_emp_raw, beta_unscaled_draws_emp, res_scaled_emp$sigma2)

# Compute posterior predictive point estimates and 95% interval bounds
pred_means_qr <- rowMeans(Y_pred_qr_emp)
pred_q025_qr  <- apply(Y_pred_qr_emp, 1, quantile, probs = 0.025)
pred_q975_qr  <- apply(Y_pred_qr_emp, 1, quantile, probs = 0.975)

# Calculate empirical 95% predictive coverage and Root Mean Squared Error (RMSE)
emp_coverage <- mean(y_emp >= pred_q025_qr & y_emp <= pred_q975_qr) * 100
emp_rmse     <- sqrt(mean((y_emp - pred_means_qr)^2))
```

#### Empirical Predictive Comparisons and Posterior Predictive Checks (PPC)

We evaluate predictive equivalence between the two samplers and verify overall model fit against observed response data $y_{\text{emp}}$:

1. **Predictive Q-Q Plots:** Compare predictive quantiles from `rawSampler` vs `rawSamplerWithQR` for representative empirical observations (Obs #1, #500, #1000).
2. **Observed vs. Predicted Plot:** Compare observed response $y_{\text{emp}}$ against posterior predictive mean $\hat{y}_{\text{pred}}$ with 95% predictive intervals.

```{r empirical-predictive-plots, fig.width=10, fig.height=8, fig.align='center'}
par(mfrow = c(2, 2), mar = c(4.5, 4.5, 2.5, 1))

# Q-Q Plots across selected empirical test observations
emp_test_indices <- c(1, 500, 1000)
invisible(lapply(emp_test_indices, function(idx) {
  qqplot(Y_pred_scaled_emp[idx, ], Y_pred_qr_emp[idx, ],
         xlab = "rawSampler Predictive Quantiles",
         ylab = "rawSamplerWithQR Predictive Quantiles",
         main = paste("Empirical Predictive Q-Q: Obs #", idx))
  abline(0, 1, col = "dodgerblue3", lwd = 2)
}))

# Posterior Predictive Check: Observed Response vs Posterior Predictive Means
plot(y_emp, pred_means_qr,
     xlab = "Observed Response (y_emp)", 
     ylab = "Posterior Predictive Mean",
     main = expression("PPC: Observed vs. Predictive Mean (" * R^2 * " Check)"),
     pch = 16, col = rgb(0.1, 0.4, 0.8, 0.3))
abline(0, 1, col = "firebrick", lwd = 2)
```

#### Predictive Performance Metrics (`data.txt`)

```{r empirical-predictive-summary, echo=FALSE}
pred_metrics <- data.frame(
  Metric = c("Posterior Predictive RMSE", "95% Credible Interval Coverage", "Max Predictive Discrepancy (QR vs Std)"),
  Value  = c(sprintf("%.4f", emp_rmse), sprintf("%.2f%%", emp_coverage), sprintf("%.2e", max(abs(rowMeans(Y_pred_qr_emp) - rowMeans(Y_pred_scaled_emp))))),
  check.names = FALSE
)

kable(pred_metrics, caption = "Empirical Posterior Predictive Metrics (data.txt)", align = "lc")
```

---

## Computational Efficiency and Execution Time Benchmarking

To evaluate the computational speed and scaling benefits of the QR reparameterization, we benchmark execution time across $N = 10,000$ draws on both simulated and empirical data using `system.time()`.

Because QR reparameterization works with $Q^\top Q = I_p$, posterior parameter sampling replaces general matrix factorizations and explicit inversions with lower/upper triangular forward- and back-substitutions (`forwardsolve()` and `backsolve()`).

```{r benchmark-timing}
# Benchmark 1: Simulated Data Runtime (n = 250, p = 3)
time_sim_standard <- system.time({
  res_sim_std <- rawSampler(y, X_scaled, m0, M0_inv, a0, b0, N_samples)
})["elapsed"]

time_sim_qr <- system.time({
  res_sim_qr <- rawSamplerWithQR(y, Q, R, m0, M0_inv, a0, b0, N_samples)
})["elapsed"]

# Benchmark 2: Empirical Data Runtime (n = 1369, p = 9)
time_emp_standard <- system.time({
  res_emp_std <- rawSampler(y_emp, X_scaled_emp, m0_emp, M0_inv_emp, a0_emp, b0_emp, N_samples)
})["elapsed"]

time_emp_qr <- system.time({
  res_emp_qr <- rawSamplerWithQR(y_emp, Q_emp, R_emp, m0_emp, M0_inv_emp, a0_emp, b0_emp, N_samples)
})["elapsed"]

# Compute relative speedup factors
speedup_sim <- time_sim_standard / time_sim_qr
speedup_emp <- time_emp_standard / time_emp_qr

# Construct summary benchmark table
benchmark_df <- data.frame(
  Dataset = c("Simulated Data (n = 250, p = 3)", "Empirical Data (n = 1369, p = 9)"),
  `Standard Sampler (s)` = sprintf("%.4f", c(time_sim_standard, time_emp_standard)),
  `QR Sampler (s)` = sprintf("%.4f", c(time_sim_qr, time_emp_qr)),
  `Speedup Factor` = sprintf("%.2fx", c(speedup_sim, speedup_emp)),
  check.names = FALSE
)

kable(benchmark_df, caption = "Execution Time Benchmarking for 10,000 Draws", align = "lccc")
```

### Key Numerical and Computational Advantages

* **Elimination of General Inversions:** Standard sampling relies on full matrix decompositions or explicit inversions of $X^\top X + M_0$. QR composition sampling operates directly on $Q$ and $R$, performing updates purely through $O(p^2)$ triangular substitutions.
* **Condition Number Optimization:** By reparameterizing $X = QR$, $Q$ has condition number $\kappa(Q) = 1$, preventing loss of precision during posterior updates even under severe predictor collinearity.
* **Vectorized Drawing Efficiency:** Generating all $N$ Monte Carlo draws in a single array operation ($Z \in \mathbb{R}^{p \times N}$) and solving $R \mathcal{B}^\top = \Theta$ via `backsolve()` completely avoids expensive R-level loop overhead.

---

## Conclusion

The primary advantage of applying the **QR decomposition method on mean-centered predictors** is that it **completely eliminates numerical ill-conditioning while preserving exact intercept anchoring around $\bar{y} \approx 11.00$**. 

By reparameterizing $X_c \beta = Q\theta$, the condition number of the orthogonalized design factor is identically $Q^\top Q = I_p = 1$, ensuring high computational speed, vectorization without loops, and total stability during precision matrix updates and posterior predictive sampling.
