---
title: "Exact Divide and Conquer Inference in Bayesian Linear Regression"
author: "Sudipto Banerjee"
output: 
  html_document:
    toc: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(dplyr)
library(knitr)
library(ggplot2)
library(tidyr)

# Source the custom divide and conquer script
source("../src/rawSamplerDivideAndConquer.r")
```

## Section 1: Partition the Data

First, we read the data, which includes the variables `Y`, `X1` through `X6`, `Lat`, and `Longi`. We partition the total 1369 rows into 10 blocks: the first 9 datasets will have exactly 130 rows, and the 10th dataset will contain the remaining 199 rows.

```{r partition_data}
# 1. Load Data
data <- read.table("../data/data.txt", header = TRUE)

# Prepare the Model Matrix and Response Vector
X_full <- as.matrix(cbind(Intercept = 1, data[, c("X1", "X2", "X3", "X4", "X5", "X6", "Lat", "Longi")]))
y_full <- as.numeric(data$Y)

# 2. Partition into Blocks
T_blocks <- 10
block_size <- 130
total_n <- nrow(X_full)

X_list <- list()
y_list <- list()

start_idx <- 1
for (t in 1:T_blocks) {
  if (t < T_blocks) {
    end_idx <- start_idx + block_size - 1
  } else {
    end_idx <- total_n # The final block absorbs the remainder (199 rows)
  }
  
  X_list[[t]] <- X_full[start_idx:end_idx, , drop = FALSE]
  y_list[[t]] <- y_full[start_idx:end_idx]
  
  start_idx <- end_idx + 1
}

cat("Successfully created", length(X_list), "data blocks.\n")
cat("Rows in block 1-9:", nrow(X_list[[1]]), "| Rows in block 10:", nrow(X_list[[10]]), "\n")
```

## Section 2: Sequential Bayesian Updating

Here we implement the block-sequential updating. The analytic expressions for the posterior at block t become the conjugate Normal-Gamma prior for block t+1. After T=10 blocks, we draw N=5000 exact samples from the final posterior distribution.

```{r seq_update}
# Setup Priors (Weakly Informative)
p <- ncol(X_full)
M0_inv <- matrix(0, nrow = p, ncol = p)
m0 <- rep(0, p)
a0 <- 0.01
b0 <- 0.01
N_samples <- 5000

# Run Divide and Conquer Sampler (T = 10)
set.seed(123)
res_T10 <- rawSamplerDivideAndConquer(
  X_list = X_list, 
  y_list = y_list, 
  M0_inv = M0_inv, m0 = m0, a0 = a0, b0 = b0, 
  N = N_samples
)

# Summarize T=10
draws_T10 <- as.data.frame(res_T10$draws)
summary_T10 <- data.frame(
  Parameter = colnames(draws_T10),
  Mean_T10 = colMeans(draws_T10),
  SD_T10 = apply(draws_T10, 2, sd)
)
kable(summary_T10, digits = 4, caption = "Posterior Summaries (Divide and Conquer, T=10)")
```

## Section 3: Comparisons

To prove the exactness of the divide and conquer approach, we now process the dataset as a single block (T=1) and compare the samples. Because the updating scheme uses sufficient statistics, the theoretical posterior is mathematically identical. 

```{r exact_comparison}
# Run Single Block Sampler (T = 1)
set.seed(123) # Use same seed to demonstrate distributional equivalence
res_T1 <- rawSamplerDivideAndConquer(
  X_list = list(X_full), 
  y_list = list(y_full), 
  M0_inv = M0_inv, m0 = m0, a0 = a0, b0 = b0, 
  N = N_samples
)

draws_T1 <- as.data.frame(res_T1$draws)
summary_T1 <- data.frame(
  Parameter = colnames(draws_T1),
  Mean_T1 = colMeans(draws_T1),
  SD_T1 = apply(draws_T1, 2, sd)
)

# Merge summaries
comparison_df <- merge(summary_T1, summary_T10, by = "Parameter")
kable(comparison_df, digits = 4, caption = "Posterior Comparison: T=1 vs T=10")
```

### Q-Q Plots for Distributional Validation

To visually confirm that the resulting distributions are practically indistinguishable, we plot the sorted posterior samples against one another. The 45-degree line represents perfect equality, and R-squared indicates the correlation.

```{r qq_plots, fig.width=10, fig.height=8}
plot_list <- list()
params_to_plot <- c("Intercept", "X1", "Lat", "sigma2")

par(mfrow=c(2,2))
for(param in params_to_plot) {
  
  # Sort samples for Q-Q alignment
  q_T1 <- sort(draws_T1[[param]])
  q_T10 <- sort(draws_T10[[param]])
  
  # Calculate R-squared for the exactness of the match
  r_squared <- cor(q_T1, q_T10)^2
  
  df_plot <- data.frame(Exact_T1 = q_T1, Sequential_T10 = q_T10)
  
  p_plot <- ggplot(df_plot, aes(x = Exact_T1, y = Sequential_T10)) +
    geom_point(alpha = 0.3, color = "steelblue") +
    geom_abline(slope = 1, intercept = 0, color = "darkred", linetype = "dashed", linewidth = 1) +
    labs(
      title = paste("Q-Q Plot:", param),
      subtitle = paste("R-squared =", round(r_squared, 6)),
      x = "Exact Posterior (T = 1)",
      y = "Sequential Posterior (T = 10)"
    ) +
    theme_bw()
  
  print(p_plot)
}
```
